---
title: "R can be your best friend"
output:
  html_document: default
  pdf_document: default
author: David Twesigomwe
date: "2026-02-26"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


## Installing/loading packages in R

### Example 1: Installing ggplot2

```{r, echo = TRUE}
install.packages("ggplot2")
```

### Loading ggplot2

```{r, echo = TRUE}
library(ggplot2)
```

### Example 2: Installing dplyr

```{r, echo = TRUE}
install.packages("dplyr")
```

### Loading dplyr

```{r, echo = TRUE}
library(dplyr)
```



## Basic data structures in R

### Variables in R

```{r, echo = TRUE}
x <- 10
y <- 20
x+y
```

### Vectors

```{r, echo = TRUE}
v <- c(1, 2, 3, 4, 5)
v
```

### Dataframe

```{r, echo = TRUE}

df <- data.frame(
  name = c("A", "B", "C"),
  age = c(23, 40, 50),
  SBP = c(111, 120, 130)
)
```

### Data exploration 

```{r, echo = TRUE}
df
```


```{r, echo = TRUE}
summary(df)
```


```{r, echo = TRUE}
dim(df)
```


```{r, echo = TRUE}
sum(df$age)
```


```{r, echo = TRUE}
mean(df$SBP)
```


## Example plot in R

### Bar plot

```{r, echo = TRUE}
barplot(df$SBP)
```


```{r, echo = TRUE}
barplot(df$SBP, main="Systolic blood pressure", xlab="Patient ID", ylab = "mmHg", names = df$name)
```


## Statistical Testing in R (demo)

### Simulate data

```{r, echo = TRUE}
BP_cases <- c(rnorm(200, mean = 140, sd = 5))
BP_controls <- c(rnorm(200, mean=120, sd = 2))
BP <- data.frame(BP_cases, BP_controls)
```

### Shapiro-Wilk test

```{r, echo = TRUE}
shapiro.test(BP_cases)
```
###  T-test (students t-test or Welch test)

```{r, echo = TRUE}
t.test(BP$BP_cases, BP$BP_controls)
```

## Importing data into R

```{r, echo = TRUE}
data <- read.csv("/Users/tdavid/Research/collaborations/IDI/rtemis/malaria_data.csv", header=TRUE, sep=",", na.strings="NA")
```


```{r, echo = TRUE}
summary(data)
```


```{r, echo = TRUE}
head(data)
```


### Simple linear regression 

```{r, echo = TRUE}
summary(lm(data$Cases ~ data$Rainfall_mm))
```

